---
title: "STAT 8830 HW5"
output: html_document
---

# 1) Local linear regression (LOESS)

Here we define a function to perform LOESS regression as a function of 1-dimensional data, a bandwidth parameter, and the number of (evenly spaced) points to predict over.

This notebook is ported more or less directly from `hw5_solutions.ipynb` and so again takes much inspiration from [this](https://github.com/joaofig/pyloess/blob/e6d12ee503fd4f924273f7cb139a10ab4e096752/pyloess/Loess.py) and [this](https://xavierbourretsicotte.github.io/loess.html).

First define some utility functions, for:

- jitter (random noise)
- point normalization
- computing band indices from bandwidth
- computing band index weights

```{r}
# set the random seed
set.seed(42)
```

# 2) Kernel density estimation

Here we define a function to perform KDE.

First define some utility functions, for:

- the kernel (normal distribution)
-

links:
- https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/bandwidth

```{r}

```
